---
title: "Alpha Diversity"
author: "Alexandra Alexiev"
date: "2025-05-14"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```

## libraries and directories

```{r libraries}
#packages and set working directory
library(vegan)
library(ggplot2)
library(mctoolsr)
library(tidyr)
library(tibble)
library(gridExtra)
library(dplyr)
library(ggpubr)
library(cowplot)
library(kableExtra)
library(RColorBrewer)
set.seed(33333)
work_dir <- "/Users/alal3493/Documents/Projects/DevelopmentalBorealToad/FungalDevData/02_PublishedAnalysis/"
seq_dir <- "/Users/alal3493/Documents/Projects/DevelopmentalBorealToad/FungalDevData/00_SeqProcessing/"

```

```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/XXX")

load("/Users/alexieva/Documents/Projects/Analysis/XXX.RData")

```

## Basic measures of highly abundant or present taxa

```{r load input files and clean up}
#load in starting files
mapfp_all <- paste0(work_dir,"Allsamps_clean_maptab.txt")
tabfp_all <- paste0(work_dir,"Allsamps_clean_otutab.txt")
input_all <- load_taxa_table(tab_fp = tabfp_all, map_fp = mapfp_all) # this file has both toad and environment sample types in it (169 samples)
mapfp_t <- paste0(work_dir,"Toad_clean_maptab.txt")
tabfp_t <- paste0(work_dir,"Toad_clean_otutab.txt")
input_toad <- load_taxa_table(tab_fp = tabfp_t, map_fp = mapfp_t) # this file only has toad samples in it (124 samples)

# order by life stage
life_list <- c("Sediment", "Water",
               "Eggs.11.15", "Tadpole.20.22",
               "Tadpole.23.25", "Tadpole.25.27",
               "Tadpole.27.29", "Tadpole.29.31",
               "Tadpole.31.35", "Tadpole.36.39",
               "Metamorph.40.46", "Subadult", "Adult")
input_all$map_loaded$Gosner_Stage <-
    factor(input_all$map_loaded$Gosner_Stage,
           levels = life_list)
input_toad$map_loaded$Gosner_Stage <-
    factor(input_toad$map_loaded$Gosner_Stage,
           levels = life_list)
# and order the other column I'll be using as well, where tadpoles are lumped
samptype_list <- c("Sediment", "Water", "Eggs",
                   "Tadpole", "Metamorph", "Subadult", "Adult")
input_all$map_loaded$Life_Stage_Simplified <-
    factor(input_all$map_loaded$Life_Stage_Simplified,
           levels = samptype_list)
input_toad$map_loaded$Life_Stage_Simplified <-
    factor(input_toad$map_loaded$Life_Stage_Simplified,
           levels = samptype_list)


```

```{r basic measures of taxa}
# number of total fungal taxa found in all samples
nrow(input_all$data_loaded) # 2156
nrow(input_toad$data_loaded) # 1760

# total number of sequences
input_sums <- input_all$data_loaded
input_sums$rowsums <- rowSums(input_sums)
colSums(input_sums) # 168627 seqs

# top ten taxa on toads and in environment
toptaxa_all <- return_top_taxa(input = input_all, number_taxa = 10)
toptaxa_toad <- return_top_taxa(input = input_toad, number_taxa = 10)
kable(toptaxa_all, caption = "Top ten taxa in toad and environment samples") %>% 
    kable_styling(bootstrap_options = c("striped","bordered"), 
                  full_width = F, position = "left")
kable(toptaxa_toad, caption = "Top ten taxa in toad samples") %>% 
    kable_styling(bootstrap_options = c("striped","bordered"), 
                  full_width = F, position = "left")


```

# Alpha diversity differences between life stages and environment samples

Below is a supplementary figure from 10.1371/journal.pone.0256328 comparing three alpha diversity measures across all the sample types. Since site is a confounding factor here, given exploratory analyses, I analyzed with blocking by site ("strata").

```{r calculate three alpha diversity metrics}
# calculate alpha diversity measures and add to metadata file as a new column
input_all$map_loaded$rich = calc_diversity(
  tax_table = input_all$data_loaded, metric = 'richness')
input_all$map_loaded$simpson = calc_diversity(
  tax_table = input_all$data_loaded, metric = 'simpson')
input_all$map_loaded$shannon = calc_diversity(
  tax_table = input_all$data_loaded, metric = 'shannon')


```

Run a Kruskal-wallis test (a non-parametric test of differences between groups) to compare alpha adversity of skin microbiome samples across life stages of the toad, with site as strata
```{r kruskal-wallis test}
# species richness
# first line runs the test; alpha diversity column is Y, metadata variables of interest are the X
krusk_rich <- kruskal.test(input_all$map_loaded$rich ~ 
                              input_all$map_loaded$Gosner_Stage)
# next, do bonferroni p value adjustment for multiple comparisons
bonpval_rich <- p.adjust(p = krusk_rich$p.value, method = "bonferroni", 
         n = length(unique(input_all$map_loaded$Gosner_Stage)))
# print values
krusk_rich; bonpval_rich

# simpson (evenness)
krusk_simp <- kruskal.test(input_all$map_loaded$simpson ~ 
                              input_all$map_loaded$Gosner_Stage)
bonpval_simp <- p.adjust(p = krusk_simp$p.value, method = "bonferroni", 
         n = length(unique(input_all$map_loaded$Gosner_Stage)))
krusk_simp; bonpval_simp

# shannon (both richness and evenness)
krusk_shan <- kruskal.test(input_all$map_loaded$shannon ~ 
                              input_all$map_loaded$Gosner_Stage)
bonpval_shan <- p.adjust(p = krusk_shan$p.value, method = "bonferroni", 
         n = length(unique(input_all$map_loaded$Gosner_Stage)))
krusk_shan; bonpval_shan


```

An alternate way to test alpha diversity differences is to use a linear model, which is especially useful if you want to test for an optimal model given many metadata variables.
```{r linear model alpha diversity test}
# step AIC is used with a linear model to test which metadata variables correlate with alpha diversity
# this is especially useful if you have multiple variables to test
# alpha diversity metric column in your metadata is the Y in the formula
# any other variables from metadata are the X
# use the metadata file as "data"
lm_shan <- lm(shannon ~ Life_Stage_Simplified + Location, 
                    data = input_all$map_loaded)
# if you have multiple variables
AIC_test <- stepAIC(lm_shan)
lm_shan <- summary(AIC_test)
lm_shan

# this should give you the results for the optimal linear model from AIC


```


```{r graph}
# Species richness graphs: mean number of OTUs found per sample type
liferich <- ggplot(input_all$map_loaded, aes(Gosner_Stage, rich)) + 
  geom_boxplot() + xlab("") + 
  ylab("Fungal OTU Richness") + ggtitle("A") +
  geom_text(x = 7, y = 194, size = 3,
            label = paste0("KW chi-sq stat=",signif(krusk_rich$statistic,
                                                    digits = 5),
                         "\n","p=",signif(bonpval_rich, digits = 3))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"),
        axis.text.y = element_text(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

# Simpson's: measures species/OTU evenness, or how close in number each species in the environment
lifesimp <- ggplot(input_all$map_loaded, aes(Gosner_Stage, simpson)) + 
  geom_boxplot() + xlab("Sample Type") + 
  ylab("Simpson's Diversity Metric") +  ylim(0,1.2) + ggtitle("B") +
  geom_text(x = 7, y = 1.15, size = 3,
            label = paste0("KW chi-sq stat=",signif(krusk_simp$statistic, 
                                                  digits = 5),
                         "\n","p=",signif(bonpval_simp, digits = 3))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"),
        axis.text.y = element_text(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

# Shannon: combination of richness and evenness
lifeshan <- ggplot(input_all$map_loaded, aes(Gosner_Stage, shannon)) + 
  geom_boxplot() + xlab("") + 
  ylab("Shannon Diversity Index") + ggtitle("C") +
  geom_text(x = 7, y = 4.1, size = 3,
            label = paste0("KW chi-sq stat=",signif(krusk_shan$statistic,
                                                  digits = 5),
                         "\n","p=",signif(bonpval_shan, digits = 3))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"),
        axis.text.y = element_text(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

alpha_panel <- grid.arrange(liferich, lifesimp, lifeshan, ncol = 3)
alpha_panel

# ggsave(paste0(work_dir,"Figs_Tables/AllAlphaDiv.png"),
#        alpha_panel, dpi = 300,
#        width = 10, height = 7)


```

# Pairwise OTU richness Stats

- Here, I'm using a two-sample Mann-Whitney test (or unpaired wilcoxon test). This is a non-parametric test of whether two populations are different across two different life stages, while still assuming those populations are independent. Since my samples came from different animals at different life stages, I need to make this assumption, so I cannot use a paired wilcoxon test.
- I calculated Bonferroni (signif if p<=0.05) and FDR (p<=0.05) p-value corrections, used both together to figure out how big of a difference there was between each life stage.

```{r pairwise significance alpha diversity}
# Are pairwise alpha diversities same or different across life stages?
nlevels(input_all$map_loaded$Gosner_Stage) # 13
cats <- levels(input_all$map_loaded$Gosner_Stage) # define category levels
cats_pairs <- combn(cats, 2) # make pairs of these
pvals <- c() # output df for pvalues
Wstat <- c() # output df for Wstat

for ( i in 1:ncol(cats_pairs) ) { # for each pair of sample types
    pair <- cats_pairs[, i] # define the pair
    Gosner_rich <- input_all$map_loaded %>% # take the mapping file
    dplyr::select(Gosner_Stage, rich) %>% # select these two columns from it
    dplyr::filter(Gosner_Stage %in% pair) # filter the sample types in the pair
    wil <- wilcox.test(rich ~ Gosner_Stage, data = Gosner_rich, paired = F, exact = F) # run Wilcoxon test
    pvals <- c(pvals, wil$p.value) # save the p val
    Wstat <- c(Wstat, wil$statistic) # save the W statistic
}
results <- data.frame(t(cats_pairs), Wstat, pvals) # make a dataframe of the useful outputs
results$pvalBon = pvals * length(pvals) # calc bonferroni pval correction
results$pvalFDR = round(pvals * (length(pvals) / rank(pvals, ties.method = "average")),
                          3) # calc FDR pval correction
results <- results %>%
  arrange(pvalBon, pvalFDR) %>% # arrange in ascending order by the corrected pvals
  filter(pvalBon <= 0.05) %>%
  filter(pvalFDR <= 0.05) %>% # filter by the two corrected pvals being <= 0.05
  dplyr::select(-pvals) # remove pval column

# write.table(results, paste0(work_dir, "Figs_Tables/pairedMW_alphadiv.txt"),
#             sep="\t", quote = F, row.names = F)

```

## save image chunk 
```{r save env, eval = F}
save.image("/Users/alexieva/Documents/Projects/Analysis/XXX.RData")

```