---
title: "Basic Beta Diversity"
author: "Alexandra Alexiev"
date: "2025-05-15"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```

## libraries and directories

```{r libraries}
# these are some commonly used ones I almost always load
library(tibble)
library(dplyr)
library(vegan)
library(ggplot2)
set.seed(3) # set seed if you're going to do certain types of stats

```

```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/XXX")

load("/Users/alexieva/Documents/Projects/Analysis/XXX.RData")

```


## Input files

```{r load input files and clean up}
#load in starting files
mapfp_all <- paste0(work_dir,"Allsamps_clean_maptab.txt")
tabfp_all <- paste0(work_dir,"Allsamps_clean_otutab.txt")
input_all <- load_taxa_table(tab_fp = tabfp_all, map_fp = mapfp_all) # this file has both toad and environment sample types in it (169 samples)
mapfp_t <- paste0(work_dir,"Toad_clean_maptab.txt")
tabfp_t <- paste0(work_dir,"Toad_clean_otutab.txt")
input_toad <- load_taxa_table(tab_fp = tabfp_t, map_fp = mapfp_t) # this file only has toad samples in it (124 samples)

# order by life stage
life_list <- c("Sediment", "Water",
               "Eggs.11.15", "Tadpole.20.22",
               "Tadpole.23.25", "Tadpole.25.27",
               "Tadpole.27.29", "Tadpole.29.31",
               "Tadpole.31.35", "Tadpole.36.39",
               "Metamorph.40.46", "Subadult", "Adult")
input_all$map_loaded$Gosner_Stage <-
    factor(input_all$map_loaded$Gosner_Stage,
           levels = life_list)
input_toad$map_loaded$Gosner_Stage <-
    factor(input_toad$map_loaded$Gosner_Stage,
           levels = life_list)
# and order the other column I'll be using as well, where tadpoles are lumped
samptype_list <- c("Sediment", "Water", "Eggs",
                   "Tadpole", "Metamorph", "Subadult", "Adult")
input_all$map_loaded$Life_Stage_Simplified <-
    factor(input_all$map_loaded$Life_Stage_Simplified,
           levels = samptype_list)
input_toad$map_loaded$Life_Stage_Simplified <-
    factor(input_toad$map_loaded$Life_Stage_Simplified,
           levels = samptype_list)


```

# Beta diversity calculation and graphing
First, calculate a beta diversity matrix. It's good practice in the exploratory phase of analysis to try 2-3 metrics, and compare them with a mantel test.
```{r calculate beta diversity matrix}
# calculate several beta diversity matrices
input_toad_rel <- convert_to_relative_abundances(input_toad)
BC <- vegdist(input_toad_rel$data_loaded, method = "bray")
EUC <- vegdist(input_toad_rel$data_loaded, method = "euclidean")
JAC <- vegdist(input_toad_rel$data_loaded, method = "jaccard")
SOR <- vegdist(input_toad_rel$data_loaded, binary = TRUE) # this is how you do sorenson distance in vegan
# another good one to try is UniFrac distance, which is applicable in R with the GUniFrac package

# partial mantel to compare if these distances are all comparable (if correlated, then yes)
mantel(BC, EUC, method = "spearman", permutations = 999)
mantel(BC, JAC, method = "spearman", permutations = 999)
mantel(BC, SOR, method = "spearman", permutations = 999)
mantel(SOR, JAC, method = "spearman", permutations = 999)
# if p value is less than 0.05, these are correlated (r tells you how correlated)
# if they are correlated, that means you can choose one without too much consideration,
# and also that it doesn't matter whether you use a distance- or dissimilarity-based metric.


```

Next, we will run a PERMANOVA to test if the groups we anticipate have different beta diversity really do. In this data set, I was wondering if 
```{r PERMANOVA}
# bray curtis matrix is Y in the formula
# input data column is X in the formula
Perm <- adonis2(BC ~ Life_Stage_Simplified, 
                data = input_toad$map_loaded,
                strata = Location)
Perm
# the R^2 in a permanova tells you what percent variation is explained by your variable(s)
# the p-value gives significance
# both of these should be reported in papers, at minimum

```

We will use the bray-curtis matrix to visualize the data in an NMDS ordination. Other types of commonly seen ordinations include PCA and PCoA.
```{r NMDS plot}
# make an NMDS plot with only the toad life stages shown
# create distance matrix
md_transformed <- t(sqrt(input_toad$data_loaded))
dm <- vegdist(md_transformed, method = "bray")

# run NMDS
md.nmds <- metaMDS(dm, k = 3, trymax = 1000) #solution reached, stress=0.198

#prepare NMDS points and values for graphing
md.nmds.points <- md.nmds$points %>% # take the NMDS points
  data.frame() %>% # make a data frame
  rownames_to_column(var = "mysamples") %>%
  mutate(SampleID = mysamples) %>%
  column_to_rownames(var = "mysamples")
meta <- input_toad$map_loaded %>%
  rownames_to_column("SampleID")
md.nmds.metadata <- inner_join(x = md.nmds.points, y = meta, 
                               by = "SampleID") %>%
  group_by(Life_Stage_Simplified) %>%
  ungroup() %>%
  dplyr::select(SampleID, MDS1, MDS2, everything()) # select these columns only
md.nmds.metadata.unq <- md.nmds.metadata %>%
  dplyr::select(Life_Stage_Simplified) %>%
  unique()

# get stress value
stress.md = paste("stress =", round(md.nmds$stress, digits = 4))
# this should be a small number, close to 0, which indicates the fit between the data and the ordination is good
# this should be reported in papers, so I have it print on the figure for ease

# get NMDS hulls
group.chulls <- plyr::ddply(md.nmds.metadata, "Life_Stage_Simplified", 
                            function(df) df[chull(df$MDS1, df$MDS2), ])

# make color pallete
samptype_cols_Toad <- c("#66A61E", 
                   "#E7298A", "#7570B3", 
                   "#D95F02", "#1B9E77")

# Plot NMDS
nmds.plot.toad <- ggplot(md.nmds.metadata, aes(x = MDS1, y = MDS2, 
                                               color = Life_Stage_Simplified)) +
  geom_polygon(data = group.chulls, 
               aes(fill = Life_Stage_Simplified), 
               alpha = 0.15, linetype = 0) +
  geom_point(size = 3) +
  annotate("text", x = Inf, y = Inf, label = stress.md, hjust = 1, vjust = 1) +
  ggtitle("D") + 
  labs(colour = "Life Stage") +
  scale_colour_manual(values = samptype_cols_Toad) +
  scale_fill_manual(values = samptype_cols_Toad) +
  guides(fill = F, color = guide_legend(order = 1, 
                                        override.aes = list(shape = 15, 
                                                            size = 6))) +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        plot.title = element_text(size = 30))

# ggsave(paste0(work_dir,"Figs_Tables/nmds.png"), 
#        nmds.plot.toad, dpi = 300,
#        width = 13, height = 8)


```

Since PERMANOVA doesn't detect differences in the dispersion of the points, we run PERMDisp next. Beta dispersion is the distance between each point in a cluster or group type and the cluster's center - we can then look at the average and variation of this dispersion. This will give us an idea of whether certain groups 
```{r test dispersion differences and make graph}
# PERMDisp on sample type
life.disper.lump <- betadisper(BC, input_all$map_loaded$Life_Stage_Simplified)
permutest(life.disper.lump)
PermDisp_life_lump <- TukeyHSD(life.disper.lump) #Tukey post-hoc
kable(PermDisp_life_lump$group) %>%
  kable_styling(bootstrap_options = c("striped","bordered"), 
                full_width = F, position = "left")

# write.table(PermDisp_life_lump$group,
#             paste0(work_dir,"Figs_Tables/PermDisp_life_lump.txt"),
#             sep = "\t")

# now make a box and whisker plot to visualize
Disper1 <- data.frame(life.disper.lump$distances)
colnames(Disper1) <- "dispers"
meta <- meta <- input_all$map_loaded %>%
  rownames_to_column("SampleID")
Disper2 <- Disper1 %>%
  rownames_to_column("SampleID") %>%
  inner_join(y = meta, 
             by = "SampleID")

disper_bars <- ggplot(data = Disper2, aes(x = Life_Stage_Simplified, 
                                          y = dispers)) +
  geom_boxplot(aes(fill = Life_Stage_Simplified), alpha = 0.6, 
               outlier.shape = NA) + 
  ylab("Dispersion") + xlab("") + ggtitle("B") +
  geom_point(aes(fill = Life_Stage_Simplified), size = 3, shape = 21, position = position_jitter(
    width = 0.25
  )) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"),
        axis.text.y = element_text(colour = "black"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        plot.title = element_text(size = 30)) +
  labs(fill = "Sample Type") +
  scale_fill_manual(values = samptype_cols)

# ggsave(paste0(work_dir,"Figs_Tables/beta_disper.png"), 
#        disper_bars, dpi = 300,
#        width = 13, height = 8)


```



## save image chunk 
```{r save env, eval = F}
save.image("/Users/alexieva/Documents/Projects/Analysis/XXX.RData")

```