---
title: "Running and graphing a basic dbRDA ordination"
author: "Alexandra Alexiev"
date: "2025-01-15"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

This code and input files are derived from this github repo: https://github.com/aalexiev/ZebrafishBaPIntergenerationalMB and from the R markdown file titled zfBaP_PubAnalysis.Rmd in that repo. This was used to publish figures such as 4C and D or 7A in https://doi.org/10.3390/toxics13010010. We used a distance-based RDA as the ordination technique because we wanted to use a model to constrain the ordination.

```{r setup, include=FALSE}
# knit options for whole doc. I will add in specific chunks when I don't want this to be the case
knitr::opts_chunk$set(echo = TRUE, # show code
                      eval = TRUE, # run chunks when knitting
                      include = TRUE, # include chunk input in final doc
                      warning = FALSE, # do not show warnings in final doc
                      message = FALSE, # do not show messages from code in final doc
                      collapse = TRUE, # when possible, do put multiple outputs in one block
                      dpi = 300, # fig resolution
                      fig.dim = c(9, 5), # the default figure dimensions
                      out.width = "98%", # the default figure output width
                      out.height = "98%", # the default figure output height 
                      cache = TRUE) # dont rerun chunks that haven't been changed

```

## libraries and directories

```{r libraries}
library(ggpubr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(gridExtra)
library(knitr)
library(stringr)
library(vegan)
library(MASS)
# set seed so stats like permanova are reproducible
set.seed(3)

```

```{r set working dir, eval=F}
setwd("/Users/alexieva/Documents/Projects/Analysis/XXX")

load("/Users/alexieva/Documents/Projects/Analysis/XXX.RData")

```

## Input data

```{r read in files}
# abundance of pathways in whole dataset
pathabund_tab_ALL <- read.csv(file = paste0(input_dir, "pathabund_tab_ALL.txt"), 
                             sep = "\t",
                             header = TRUE)

# metadata for whole dataset
metaALL <- read.csv(file = paste0(input_dir, "meta_tab_ALL.txt"),
                   sep = "\t",
                   header = TRUE) %>%
  mutate(Exposure = relevel(factor(Exposure), "DMSO"), Morpholino = relevel(factor(Morpholino), "CoMo"),
         Generation = factor(Generation),
         Sex = factor(Sex), # relevel and factor so these are accurate
         BMI = Weight_g / (Length_mm/1000)^2) # add BMI calculation

# reorder factors for treatment
metaALL$Treatments <- factor(x = metaALL$Treatments,
                             levels = c("CoMo_DMSO",
                                        "CoMo_BaP",
                                        "AHR2Mo_DMSO",
                                        "AHR2Mo_BaP"))


```

## Select optimal model
Use ordistep to select the model that best explains the data in a dbRDA. You can use ordistep with other types of constrained ordination in vegan. Capscale() is the function that corresponds with a dbRDA.
```{r ordistep model selection, eval=F}
# first subset the data to include only the relevant variables
meta_modALL <- metaALL %>%
  dplyr::select(c("Exposure", "Morpholino", "Generation", "Sex", "BMI"))
# make a model construct with the intercept only; this is to see if the models with variables are actually a better fit compared to...well...no variables.
ALLmod0 <- capscale(pathabund_tab_ALL ~ 1, meta_modALL, distance = "bray")
# make a model construct with all variables in the column of the metadata file (the period) and a few interactions we were interested in (the colon variables)
ALLmod1 <- capscale(pathabund_tab_ALL ~ . + Exposure:Morpholino + Morpholino:Exposure + Generation:Exposure + Generation:Morpholino + Sex:BMI, meta_modALL, distance = "bray")
# this determines what the best model is to run RDA on
pw_ordiALL <- ordistep(ALLmod0, scope = formula(ALLmod1))
pw_ordiALL

# best model formula is listed at the top
# model inertia is scaled Chi-square, worth recording
# proportion variance explained is also given and worth recording

```

Now run the dbRDA with the formula from the selected optimal model determined by ordistep
```{r run dbRDA}
# calculate RDA (distance based), using optimal model formula from ordistep above
pw_rda <- capscale(formula = pathabund_tab_ALL ~ Generation + Exposure + Morpholino + Generation:Exposure + Exposure:Morpholino + Generation:Morpholino, 
                   data = meta_modALL, distance = "bray")
# using bray-curtis distance again
# have to adjust the R squared manually
RsquareAdj(pw_rda)
# this gives the amount of variation explained by the model


```

# Graph the ordination
```{r graph dbRDA}
# create input files for graphing
smry_rda <- summary(pw_rda)

# this is the coordinates for the points and metadata
pw_PC1  <- data.frame(smry_rda$sites[,1:2]) %>%  # these are the x, y coordinates for the sample points
  rownames_to_column("SampleID") %>%
  inner_join(dplyr::select(metaALL, c(SampleID, Treatments, Generation, Exposure, Morpholino)), by = "SampleID") %>%
  mutate(treatgen = paste0(Treatments, "_", Generation)) %>%
  column_to_rownames("SampleID")

# reorder the treatments
pw_PC1$Treatments <- factor(x = pw_PC1$Treatments,
                             levels = c("CoMo_DMSO",
                                        "CoMo_BaP",
                                        "AHR2Mo_DMSO",
                                        "AHR2Mo_BaP"))
# set the graph axis limits
mylims <- range(with(pw_PC1, c(CAP1, CAP2)))

# the biplot scores are the correlations between your environmental variables and axes
pw_PC2  <- data.frame(smry_rda$biplot)

# put the plot together in ggplot
pwrda_plot <- ggplot(pw_PC1, aes(x = CAP1, y = CAP2)) + 
  geom_point(aes(color = Treatments), size = 2) +
  stat_ellipse(aes(group = treatgen, 
                   color = Treatments), 
               show.legend = T) +
  theme_classic() +
  labs(x = paste0("CAP1 (",round(100*smry_rda$cont$importance[2, "CAP1"], digits = 2),"%)"), # the importance is the variation explained by each axis
       y = paste0("CAP2 (",round(100*smry_rda$cont$importance[2, "CAP2"], digits = 2),"%)")) +
  scale_color_brewer(palette = "PuOr", direction = -1, # this direction argument just reverses the order of colors in the pre-made palette
                     labels = c("AhR2Mo - / BaP -",
                               "AhR2Mo - / BaP +",
                               "AhR2Mo + / BaP -",
                               "AhR2Mo + / BaP +")) +
  theme(text = element_text(size = 20)) +
  scale_shape_discrete(name = "Generation shape") +
  scale_linetype_discrete(name = "Ellipse line type") +
  facet_wrap("Generation")

pwrda_plot

```

# Stats
Lastly, we run a PERMANOVA on the optimal model formula from ordistep above
```{r permanova best model}
# calculate bray curtis matrix on an abundance matrix
pw_dm <- vegdist(pathabund_tab_ALL, method = "bray")

# run a PERMANOVA where X is variables from the optimal model formula from above
# Y is the bray-curtis matrix
PermExpandMod_pw <- adonis2(pw_dm ~ Generation + Exposure + Morpholino + Generation:Exposure + Exposure:Morpholino + Generation:Morpholino, data = metaALL)

PermExpandMod_pw
# record the R^2 (amount of variation explained by each variable tested)
# record p-value 

```


## save image chunk 
```{r save env, eval = F}
save.image("/Users/alexieva/Documents/Projects/Analysis/XXX.RData")

```